---
title: "Bayesian Marginal Reconstruction"
author: "Nathan Morris"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing 'tboot_bmr' vs. Joint Bayesian Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


#Comparing Bayesian Marginal Reconstruction (BMR) with Full Multivariate Bayesian Analysis for Multivariate Binary Data

The following example is conceived to demonstrate the similarity of BMR to full joint Bayesian analysis in some cases.

##The Example Scenario and Simulated Datasets
First, we create several scenarios where we assume we have 4 variables each of which is binary leading to $2^4=16$ possible outcomes in combination. We assign probabilities to each combination for 2 different simulated scenarios.

```{r cache = FALSE}
library(tboot)
library(ggplot2)

possible_values=as.matrix(expand.grid(x1=c(0,1),x2=c(0,1),x3=c(0,1),x4=c(0,1)))
head(possible_values)

normalize=function(x) x/sum(x)

#sceanario 1 assigns some probability to each possible_value
M=rbind(c(-2.7, 0, 0, 0),
c(4.05, -2.7, 0, 0),
c(-0.13, 3.48, -3.13, 0),
c(1.49, -2.12, 2.9, -1.13))


prob_1=normalize(exp( apply(possible_values, 1, 
                            function(x) x%*%M%*%x))) 

#sceanario 2 exontially tilts scenario 1
prob_2=normalize(prob_1*exp( apply(possible_values, 1, 
                            function(x) .5*sum(x)))) 

```

To understand the probabilities above consider the following correlation and mean for the binary variables being simulated:

```{r cache = FALSE}

scenarioCorrelation=function(wt) {
  mycov=cov.wt(possible_values,wt)
  return(cov2cor(mycov$cov))
}
scenarioMean=function(w) 
  t(crossprod(possible_values,w))

round(scenarioCorrelation(prob_1),2)
round(scenarioMean(prob_1),2)
round(scenarioCorrelation(prob_2),2)
round(scenarioMean(prob_2),2)
```

Next we simulate data sets with a range of sample sizes. For each possible sample size we simulate 50 datasets.

```{r cache = FALSE}
sim = function(N_per_datset,N_datasets, p) {
  replicate(N_datasets, {
    index=sample.int(nrow(possible_values), N_per_datset,replace = TRUE, prob=p)
    possible_values[index,]
  }, simplify = FALSE)
}

ss=c(25, 50, 200, 400)
sim_data_1= lapply(ss, function(N_per_datset) 
  sim(N_per_datset,50,prob_1))
sim_data_2= lapply(ss, function(N_per_datset) 
  sim(N_per_datset,50,prob_2))
names(sim_data_1)=names(sim_data_2)=paste0("N=",ss)
```

##Comparing BMR Posterior Distributions from post_bmr vs Full Bayesian Analysis

Using the above simulated datasets we use marginal Bayesian calculations followed by BMR to sample from the joint posterior. We assume a simple uniform prior for the marginal simulations.

```{r cache = FALSE}
set.seed(2020)


#this function doest the actual analysis
get_post_bmr= function(marginal_data,individual_data) {
  marginal=lapply(data.frame(marginal_data),
                  function(x) 
                    rbeta(5000,sum(x)+1, sum(x==0) + 1))
  w=tweights_bmr(individual_data, marginal,
                 silent = TRUE, warningcut = .1)
  return(post_bmr(5000, w))
}


#If twieights or tboot throw an error or warning assume 
# (i.e. BMR analysis should not be performed in these cases)
missingSim=matrix(NA,5000,4); colnames(missingSim)=c("x1","x2","x3","x4")

#analyze all simulated datasets
anayze_all_BMR=function(sim_dta, sim_dta_cor) {
  mapply(function(s1a,s2a) { #loop over sample sizes
    mapply(function(s1b,s2b) { #loop over simulations
      tryCatch(get_post_bmr(s2b, s2b), 
               error=function(e) return(missingSim),
               warning=function(w) return(missingSim))
    },s1a, s1a, SIMPLIFY=FALSE)
  },sim_dta, sim_dta_cor, SIMPLIFY=FALSE)
}

#infer correlation from the same dataset
bmr_results_1=anayze_all_BMR(sim_data_1,sim_data_1)

#infer the correlation from a different dataset
bmr_results_2=anayze_all_BMR(sim_data_1,sim_data_2)


```
Next we analyze these datasets by using a fully Bayesian analysis. We consider that
the prior probabilities assigned to each possible outcome (i.e., 2^4 possible outcomes) follow a Dirichlet prior with parameter rep(1/8,16). This prior is marginally uniform for each variable.

```{r cache = FALSE}

rdirichlet  = function(n, alpha) {
  x <- matrix(rgamma(n*length(alpha), alpha), nrow=length(alpha))  
  return( t(x) / colSums(x)) 
}

joint_bayes=function(dta) {
  dta_factor=factor(apply(dta, 1, function(x) paste0(x, collapse = ",")),
                    levels = apply(possible_values, 
                                   1, function(x) paste0(x, collapse = ",")))
  tbl=table(dta_factor)
  post_prob=rdirichlet(5000, as.vector(tbl)+2/(2^4))
  prob_proportions=post_prob %*% possible_values
  return(prob_proportions)
}

anayze_all_joint=function(sim_dta, sim_dta_cor) {
  lapply(sim_dta, function(sa) { #loop over sample sizes
    lapply(sa, function(sb) { #loop over simulations
      return(joint_bayes(sb))
    })
  })
}

joint_results=anayze_all_joint(sim_data_1)

```

Next we set up several functions to graphically compare the two methods. We will use a series of Bland–Altman plots to compare properties of the distributions. In all of these graphs, each dot represents a posterior values (such as median, correlation or quantile) from a simulated sample of size N. 

```{r cache = FALSE}

#Extract any summary measure of interest for the data.
# g is the function that specifies which property to extract
get_property=function(g) {
  ret=do.call(rbind, mapply(function(s1a,s2a,s3a, N) { #loop over sample sizes
    ret=do.call(rbind,mapply(function(s1b,s2b, s3b) { #loop over simulations
      J=g(s3b)
      b1=g(s1b)
      b2=g(s2b)
      return(data.frame(N=N, 
                        analysis=c("Marginals Generated\nfrom Individual Data", 
                                   "Marginal Generated\nfrom Seperate Data"),
                        dif=c(b1-J, b2-J),mean=c((b1+J)/2, (b2+J)/2)))
    },s1a,s2a,s3a, SIMPLIFY = FALSE))
    ret$sim_id=rep(1:(nrow(ret)/2), each=2)
    return(ret)
  },bmr_results_1, bmr_results_2, joint_results, names(bmr_results_1), SIMPLIFY = FALSE))
  rownames(ret)=NULL
  ret=na.omit(ret)
  return(ret)
}

make_graph=function(g,title) {
  pltdta=get_property(g)
  ggplot(pltdta, aes(y=dif, x=mean, color=N)) + 
    geom_point() + ggtitle(title) + facet_grid(~analysis) +
    geom_hline(yintercept=0) + ylab("Difference(BMR-Joint)") + xlab("Mean (BMR+Joint)")
}

```

Now we investigate the marginals graphically. As may be seen the quantiles and median of the joint method and BMR are very similar. The similarity grows stronger as the sample size increases.

```{r cache = FALSE}
for(i in 1:4)
  show(make_graph(function(d) median(d[,i]), paste0("Median of X",i)))

for(i in 1:4)
  show(make_graph(function(d) quantile(d[,i], .05, na.rm=TRUE),
                  paste0("5% Quantile of X",i)))
for(i in 1:4)
  show(make_graph(function(d) quantile(d[,i], .95, na.rm=TRUE),
                  paste0("95% Quantile of X",i)))
```

Similarly we investigate the correlation between several variables. We will not print out all the correlations. 

```{r cache = FALSE}
for(i in 1:3)
  for(j in (i+1):4)
  show(make_graph(function(d) cor(d[,i],d[,j]),
                  paste0("Correlation of X",i, " and X",j )))
```


Finally we investigate the quantiles in several random directional transformations of the space:
```{r cache = FALSE}
for(i in 1:5) {
  direction=rnorm(4)
  direction=direction/sum(direction^2)
  show(make_graph(
    function(d) {
      quantile(d %*% direction, .95, na.rm=TRUE)},
    paste0("Quantile of p %*% c(",paste(round(direction,2), collapse=","),")")
  )
  )
}

  
```



Finally we investigate the quantiles in several random directional transformations of the space:
```{r cache = FALSE}
for(i in 1:5) {
  direction=rnorm(4)
  direction=direction/sum(direction^2)
  show(make_graph(
    function(d) {
      quantile(d %*% direction, .95, na.rm=TRUE)},
    paste0("Quantile of p %*% c(",paste(round(direction,2), collapse=","),")")
  )
  )
}
  
```


##Comparing 'tboot_bmr' vs Full Bayesian Analysis
Consider that we want to test the hypothesis for each variable that the proportion is 0.6. We wish to calculate the probability of succeeding on all 4 endpoints with a future trial sample size of 400. Also, consider that the data available comes from a sample size of 50 patients as previously simulated. We will only look at 10 simulated datasets because the simulation is computationally expensive with 1000 future simulated trials based on each dataset.

```{r cache = FALSE}
set.seed(2020)

sim_data_1_50= sim_data_1[["N=50"]][1:10]
sim_data_2_50= sim_data_2[["N=50"]][1:10]


#this function simulates a series of trials with success or failure
#for each trial using tboot_bmr
get_trial_pobS_bmr= function(marginal_data,individual_data) {
  marginal=lapply(data.frame(marginal_data),
                  function(x) 
                    rbeta(5000,sum(x)+1, sum(x==0) + 1))

  w=tweights_bmr(individual_data, marginal,
                 silent = TRUE, warningcut = .15)
  S=t(replicate(1000, {
    dta=tboot_bmr(400, w)
    p=colMeans(dta)
    se=sqrt(p*(1-p)/nrow(dta))
    abs(p-.6)/se>qnorm(0.975) #success for a 2-sided alpha 0.05
  }))
  #return the proportion of time we succeed on all endpoints
  return(mean(apply(S,1,all))) 
}


#this function simulates a series of trials with success or failure
#for each trial using full Bayes analysis
get_trial_probS_joint= function(dta) {
  dta_factor=factor(apply(dta, 1, function(x) paste0(x, collapse = ",")),
                    levels = apply(possible_values, 
                                   1, function(x) paste0(x, collapse = ",")),
                    ordered = TRUE)
  tbl=table(dta_factor)
  post_prob=rdirichlet(1000, as.vector(tbl)+2/(2^4))

  S=t(apply(post_prob, 1, function(p) {
    index=sample.int(nrow(possible_values), 400,replace = TRUE, prob=p)
    dta=possible_values[index,]
    p=colMeans(dta)
    se=sqrt(p*(1-p)/nrow(dta))
    abs(p-.6)/se>qnorm(0.975) #success 
  }))
  #return the proportion of time we succeed on all endpoints
  return(mean(apply(S,1,all))) 
}




#infer correlation from the same dataset
bmr_probS_1=mapply(get_trial_pobS_bmr, sim_data_1_50, sim_data_1_50)

#infer the correlation from a different dataset
bmr_probS_2=mapply(get_trial_pobS_bmr, sim_data_1_50, sim_data_2_50)

#perform full bayesian analysis
joint_probS=sapply(sim_data_1_50, get_trial_probS_joint)
  
```

As may be seen below, the 3 methods yield very similar results.
```{r cache = FALSE}
print(data.frame(bmr_probS_1,bmr_probS_2,joint_probS))

```

